Creating Data for "Social inequality in mortality among adults and elderly in northern Sweden 1850--2013"\footnote{Paper presented at CEDAR, 27 April 2017, Version 2017-03-28.}

### 2017-05-18 CreateData/creDat.Rnw
fix typo
refresh

\title{Creating Data for "Social inequality in mortality among
  adults and elderly in northern Sweden 1850--2013"\footnote{Paper presented
    at the
    session 'Socoeconomic inequalities and mortality I: Europe'. 41st Annual Meeting of the Social Science History Association, 
    Chicago, Illinois
    November 17--20, 2016. Version 2016-09-19.}}

\date{November 19, 2016}

@@

\begin{abstract}
This version is 'old', but all the code is included, so it will be crystal-clear to the reader exactly what has been done.

Spotted errors must pe reported immediately!

@@

@@

\begin{enumerate}
\item \'Elite, HISCLASS 1 and 2, Higher and lower managers, and higher
  professionals. 
\item HISCLASS 3  and 8
\item HISCLASS 4, 5, 6. Farmers, fishermen, and lower professionals, lower clerical, foremen.
\item Workers, HISCLASS 7, 9, 10, 11, and 12.
%%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
\end{enumerate}

@@

<<getdata>>=
source("calWin.R")
calWin
library(xtable)
library(knitr)
library(eha)
library(skum)
library(dplyr)
library(tidyr)
library(stringr)
load("../../../Data/adult.rda")
adult <- calWin(adult, c(1851, 1951))
source("R/addPer.R")
addPer
adult <- addPer(adult, cuts = seq(1851, 1951, by = 25))
adult <- rc(adult)
## NEW hisclass, 6 + 7 --> 6:
##adult$hisclass[!is.na(adult$hisclass) & (adult$hisclass == 7)] <- 6
## Recode to four classes:
##New Hisclasses:
ett <- with(adult, !is.na(socBranch) &
                ((socBranch == "official" & socStatus == "high") |
                     (socBranch == "business" & socStatus == "high")
                ))
tva <- with(adult, !is.na(socBranch) &
                ((socBranch == "farming" & socStatus == "high") |
                     (socBranch == "business" & socStatus == "low")
                ))
tre <- with(adult, !is.na(socBranch) &
                    (socBranch == "official" & socStatus == "low")
               )
fyra <- with(adult, !is.na(socBranch) &
                ((socBranch == "farming" & socStatus == "low") |
                     (socBranch == "worker"))
                )

hisc <- rep(-1, NROW(adult))
hisc[ett] <- 1
hisc[tva] <- 2
hisc[tre] <- 3
hisc[fyra] <- 4
is.na(hisc) <- hisc == -1
adult$hisclass <- hisc 
# hisc <- adult$hisclass
#     hisc[is.na(hisc)] <- -1
#     x <- hisc
#     hisc[x %in% 1:2] <- 1
#     hisc[x %in% 3:4] <- 2
#     hisc[x == 5] <- 3
#     hisc[x >= 5.9] <- 4
#     is.na(hisc) <- hisc == -1
#     adult$hisclass <- hisc

@@

<<ssykonfob>>=
    ##load("../../../Data/fobdata.rda")
    load("data4/fobdata.rda")
    ## Cause of death:
    load("../../../Data/socstyr.rda")
    indx <- match(fobdata$LINNEID, socstyr$LINNEID)
    fobdata$ULORSAK <- str_trim(as.character(socstyr$ULORSAK[indx]))
    source("R/putONssyk.R")
    putONssyk
    fobb <- putONssyk(fobdata)
    source("R/ssykTOhisclass.R")
    ssykTOhisclass
    fobb$hisclass <- ssykTOhisclass(fobb$ssyk)
    is.na(fobb$hisclass) <- with(fobb, is.na(hisclass) |
                                       (!is.na(hisclass) & hisclass == -1))
    ##source("R/pushHis.R")
    ##fobb <- pushHis(fobb)
    ##fobb$hisclass <- fobb$nyhisclass
    ## Instead
    fobb <- group_by(fobb, LINNEID) %>% fill_("hisclass")
    fobb$id <- fobb$LINNEID
    fobb <- addPer(fobb, cuts = c(1961, 1971, 1981, 1991, 2002))
    load("../../../Data/sex.rda")

@@

    ##load("../../../Data/lisadata.rda")
    load("data4/lisadata.rda")
    ## Cause of death:
    load("../../../Data/socstyr.rda")
    indx <- match(lisadata$LINNEID, socstyr$LINNEID)
    lisadata$ULORSAK <- str_trim(as.character(socstyr$ULORSAK[indx]))
    civst <- lisadata$CIV
    civst[is.na(civst)] <- 0
    civst[civst == 4] <- 3
    civst[civst == 5] <- 2
    civst[civst == 6] <- 3
    is.na(civst) <- civst == 0
    lisadata$civst <- civst
    rm(civst)
    ## Fix KON:
    load("../../../Data/sex.rda")
    indx <- match(lisadata$LINNEID, sex$LINNEID)
    lisadata$sex <- sex$KON[indx]
    levels(lisadata$sex) <- c("male", "female")
    lisaold <- lisadata
    lisadata <- calWin(lisadata, c(2002, 2014))

@@

    q()
    ## This is new: Do not remove:
    ##lisadata <- lisadata[!is.na(lisadata$SSYK3), ] ##NOTE!

@@

vb <- vb[order(vb$id, vb$age), ]
source("../R/getCancer.R")
getCancer
vb <- getCancer(vb)
source("../R/getHeart.R")
getHeart
vb <- getHeart(vb)
vb$exposure <- with(vb, exit - enter)
vb0 <- vb
vb <- vb[!is.na(vb$hisclass), ] # same as 'na.omit(vb)' (I think)
first <- function(x) x[1]
hisclass40 <- with(vb, tapply(hisclass, id, first))
indx <- match(vb$id, names(hisclass40))
vb$hisclass40 <- hisclass40[indx]
##
vb$hisclass60 <- 0
hisclass60 <- with(vb[vb$age > 59, ], tapply(hisclass, id, first))
indx <- match(vb$id, names(hisclass60))
vb$hisclass60 <- hisclass60[indx]
komp <- is.na(vb$hisclass60)
vb$hisclass60[komp] <- vb$hisclass40[komp]
vb$hisclass40 <- factor(vb$hisclass40, labels = c("1+2", "3+8", "4+5+6", "7+9+10+11+12"))
vb$hisclass60 <- factor(vb$hisclass60, labels = c("1+2", "3+8", "4+5+6", "7+9+10+11+12"))
vb$hisclass <- factor(vb$hisclass, labels = c("1+2", "3+8", "4+5+6", "7+9+10+11+12"))
## End cut ####################################################
levels(vb$period)[8] <- "1991-2001" ## Changed from "1991-2000", fob90 extended.
levels(vb0$period)[8] <- "1991-2001"
save(vb0, file = "../data/vb0.rda")
save(vb, file = "../data/vb.rda")
@

@@

<<aggregate>>=
vbTab <- aggregate(vb[, c("event", "exposure")], 
                   vb[, c("sex", "age", "period", "cohort", 
                          "urban", "civst", "hisclass40", "hisclass60")], 
                   FUN = sum)
## Is this necessary: (See above!)
##vbTab$hisclass40 <- factor(vbTab$hisclass40)
##vbTab$hisclass60 <- factor(vbTab$hisclass60)
vbTab$age <- factor(vbTab$age)
save(vbTab, file = "../data/vbTab.rda")
@

Thus, the one data file to use is {\tt vbTab.rda}.

<<lisaold>>=
lisaold <- age.window(lisaold, c(40, 90))
lisaold$exposure <- with(lisaold, exit - enter)
lisaold$educ <- as.factor(lisaold$SUN2000NIVA_OLD)
lisaold$income <- cut(lisaold$DISPINKPERSF, c(-600, 730, 958, 1270, 500000),
                      labels = c("lowest", "low", "high", "highest"))
loTab <- aggregate(lisaold[, c("event", "exposure")], 
                   lisaold[, c("sex", "age", "period", "urban", 
                               "civst", "income", "educ")],
                   FUN = sum)
save(loTab, file = "../data/loTab.rda")
@

>>>>

\title{Creating Data for "Social inequality in mortality among
  adults and elderly in northern Sweden 1850--2013"\footnote{Paper presented
    at CEDAR, 27 April 2017,
    Version 2017-03-28.}}

\date{March 28, 2017}

@@

\begin{abstract}
This version is 'new', but all the code is included, so it will be crystal-clear to the reader exactly what has been done.

Spotted errors must be reported immediately!

@@

In this version the hisclasses are grouped into three levels instead of four.

@@

\begin{enumerate}
\item {\bf elite}, HISCLASS 1 and 2, Higher and lower managers, and higher
  professionals. 
%%\item HISCLASS 3  and 8
%%\item HISCLASS 4, 5, 6. Farmers, fishermen, and lower professionals, lower clerical, foremen.
\item {\bf middle} HISCLASS 3, 4, 5, 6, and 8. Farmers, fishermen, and lower professionals, lower clerical, foremen.
\item {\bf worker}, HISCLASS 7, 9, 10, 11, and 12.
%%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
\end{enumerate}

@@

<<getdata>>=
source("calWin.R")
calWin
library(xtable)
library(knitr)
library(eha)
library(skum)
library(dplyr)
library(tidyr)
library(stringr)
load("../../Data/adult.rda")
adult <- calWin(adult, c(1851, 1951))
source("R/addPer.R")
addPer
adult <- addPer(adult, cuts = seq(1851, 1951, by = 25))
adult <- rc(adult)
## NEW hisclass, 6 + 7 --> 6:
##adult$hisclass[!is.na(adult$hisclass) & (adult$hisclass == 7)] <- 6
## Recode to four classes:
##New Hisclasses:
ett <- with(adult, !is.na(socBranch) &
                ((socBranch == "official" & socStatus == "high") |
                     (socBranch == "business" & socStatus == "high")
                ))
##tva <- with(adult, !is.na(socBranch) &
  ##              ((socBranch == "farming" & socStatus == "high") |
##                     (socBranch == "business" & socStatus == "low")
##                ))
##tre <- with(adult, !is.na(socBranch) &
##                    (socBranch == "official" & socStatus == "low")
##               )
tva <- with(adult, !is.na(socBranch) &
                ((socBranch == "farming" & socStatus == "high") |
                 (socBranch == "business" & socStatus == "low") |
                 (socBranch == "official" & socStatus == "low")
                ))

tre <- with(adult, !is.na(socBranch) & ## Original 'fyra' ---> 'tre'
                ((socBranch == "farming" & socStatus == "low") |
                     (socBranch == "worker"))
                )

hisc <- rep(-1, NROW(adult))
hisc[ett] <- 1
hisc[tva] <- 2
hisc[tre] <- 3
##hisc[fyra] <- 4
is.na(hisc) <- hisc == -1
adult$hisclass <- hisc 
# hisc <- adult$hisclass
#     hisc[is.na(hisc)] <- -1
#     x <- hisc
#     hisc[x %in% 1:2] <- 1
#     hisc[x %in% 3:4] <- 2
#     hisc[x == 5] <- 3
#     hisc[x >= 5.9] <- 4
#     is.na(hisc) <- hisc == -1
#     adult$hisclass <- hisc

@@

<<ssykonfob>>=
    ##load("../../Data/fobdata.rda")
    load("data4/fobdata.rda")
    ## Cause of death:
    load("../../Data/socstyr.rda")
    indx <- match(fobdata$LINNEID, socstyr$LINNEID)
    fobdata$ULORSAK <- str_trim(as.character(socstyr$ULORSAK[indx]))
    source("R/putONssyk.R")
    putONssyk
    fobb <- putONssyk(fobdata)
    source("R/ssykTOhisclass.R")
    ssykTOhisclass
    fobb$hisclass <- ssykTOhisclass(fobb$ssyk)
    is.na(fobb$hisclass) <- with(fobb, is.na(hisclass) |
                                       (!is.na(hisclass) & hisclass == -1))
    ##source("R/pushHis.R")
    ##fobb <- pushHis(fobb)
    ##fobb$hisclass <- fobb$nyhisclass
    ## Instead
    fobb <- group_by(fobb, LINNEID) %>% fill_("hisclass")
    fobb$id <- fobb$LINNEID
    fobb <- addPer(fobb, cuts = c(1961, 1971, 1981, 1991, 2002))
    load("../../Data/sex.rda")

@@

    ##load("../../../Data/lisadata.rda")
    load("data4/lisadata.rda")
    ## Cause of death:
    load("../../Data/socstyr.rda")
    indx <- match(lisadata$LINNEID, socstyr$LINNEID)
    lisadata$ULORSAK <- str_trim(as.character(socstyr$ULORSAK[indx]))
    civst <- lisadata$CIV
    civst[is.na(civst)] <- 0
    civst[civst == 4] <- 3
    civst[civst == 5] <- 2
    civst[civst == 6] <- 3
    is.na(civst) <- civst == 0
    lisadata$civst <- civst
    rm(civst)
    ## Fix KON:
    load("../../Data/sex.rda")
    indx <- match(lisadata$LINNEID, sex$LINNEID)
    lisadata$sex <- sex$KON[indx]
    levels(lisadata$sex) <- c("male", "female")
    lisaold <- lisadata
    lisadata <- calWin(lisadata, c(2002, 2014))

@@

    ## This is new: Do not remove:
    ##lisadata <- lisadata[!is.na(lisadata$SSYK3), ] ##NOTE!

@@

vb <- vb[order(vb$id, vb$age), ]
source("../R/getCancer.R")
getCancer
vb <- getCancer(vb)
source("../R/getHeart.R")
getHeart
vb <- getHeart(vb)
vb$exposure <- with(vb, exit - enter)
vb0 <- vb
vb <- vb[!is.na(vb$hisclass), ] # same as 'na.omit(vb)' (I think)
first <- function(x) x[1]
hisclass40 <- with(vb, tapply(hisclass, id, first))
indx <- match(vb$id, names(hisclass40))
vb$hisclass40 <- hisclass40[indx]
##
vb$hisclass60 <- 0
hisclass60 <- with(vb[vb$age > 59, ], tapply(hisclass, id, first))
indx <- match(vb$id, names(hisclass60))
vb$hisclass60 <- hisclass60[indx]
komp <- is.na(vb$hisclass60)
vb$hisclass60[komp] <- vb$hisclass40[komp]
vb$hisclass40 <- factor(vb$hisclass40, labels = c("1+2", "3+4+5+6+8", "7+9+10+11+12"))
vb$hisclass60 <- factor(vb$hisclass60, labels = c("1+2", "3+4+5+6+8", "7+9+10+11+12"))
vb$hisclass <- factor(vb$hisclass, labels = c("1+2", "3+4+5+6+8", "7+9+10+11+12"))
## End cut ####################################################
levels(vb$period)[8] <- "1991-2001" ## Changed from "1991-2000", fob90 extended.
levels(vb0$period)[8] <- "1991-2001"
save(vb0, file = "../data/vb0.rda")
save(vb, file = "../data/vb.rda")
@

@@

<<aggregate>>=
vb$poplink <- vb$dataset == "poplink"
vbTab <- aggregate(vb[, c("event", "exposure")], 
                   vb[, c("sex", "age", "period", "cohort", 
                          "urban", "civst", "hisclass40", "hisclass60", "poplink")], 
                   FUN = sum)

vb0$hisclass[is.na(vb0$hisclass)] <- 0
vb0$poplink <- vb0$dataset == "poplink"
vb0Tab <- aggregate(vb0[, c("event", "exposure")], 
                   vb0[, c("sex", "age", "period", "cohort", 
                          "urban", "civst", "hisclass", "poplink")], 
                   FUN = sum)
## Is this necessary: (See above!)
##vbTab$hisclass40 <- factor(vbTab$hisclass40)
##vbTab$hisclass60 <- factor(vbTab$hisclass60)
vbTab$age <- factor(vbTab$age)
save(vbTab, file = "../data/vbTab.rda")
save(vb0Tab, file = "../data/vb0Tab.rda")
@

Thus, the one data file to use is {\tt vbTab.rda} or {\tt vb0Tab.rda}.

<<lisaold>>=
lisaold <- age.window(lisaold, c(40, 90))
lisaold$exposure <- with(lisaold, exit - enter)
lisaold$educ <- as.factor(lisaold$SUN2000NIVA_OLD)
lisaold$income <- cut(lisaold$DISPINKPERSF, c(-600, 730, 958, 1270, 500000),
                      labels = c("lowest", "low", "high", "highest"))
loTab <- aggregate(lisaold[, c("event", "exposure")], 
                   lisaold[, c("sex", "age", "period", "urban", 
                               "civst", "income", "educ")],
                   FUN = sum)
save(loTab, file = "../data/loTab.rda")
save(lisaold, file = "../data/lisaold.rda")
@
